An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems

### 2014-04-12 paper/paper.tex
Paper: update background sections 'P2P Backup Systems' and 'Cryptography', among other small changes

%%\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{alltt}
\usepackage{color}
\usepackage{fancyvrb} 
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{url}
\usepackage[autostyle]{csquotes} 
\usepackage[toc,page]{appendix}

@@

\begin{document}
\begin{center}
\thispagestyle{empty}
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\newpage
\end{center}

@@

In the client-server model, servers are expected to be always available and accessible. Peer to peer networks do not have this luxury. Instead, they experience the constant fluctuation of peers joining and leaving the network. This is called \textit{churn}. Churn in P2P networks complicates the communication between peers because the sought peer may be offline. This is especially a problem in P2P backup systems, for example, when trying to recover data from a peer who is not online. The strategies for dealing with churn, and many other P2P-specific complications, heavily depend on the type of overlay network a P2P application uses. We discuss churn in more detail in section \ref{sec:churn}.

@@

An overlay network is a network built on top of another network. That is, an overlay network consists of nodes with neighbors who are not physically connected, but instead use the underlying network to establish a virtual connection. Consider a trivial example where there are three computers connected to the Internet, but located in different parts of the world. These three computers could form an overlay network by agreeing to be virtual neighbors to each other. If their overlay network was used in a P2P application, then, when connecting to each other, they will use the underlying network, the Internet, to actually send messages. The P2P application does not need to know about the underlying network that is used to route messages, it only needs to know that there is an overlay network that defines the direct connections between peers.

As alluded to, peer to peer networks are implemented using overlay networks on top of existing infrastructure.\footnote{Overlay networks are used for many purposes, not just P2P applications. Note, however, in discussing overlay networks, we will refer to their participants as \textit{peers}.} This allows peers to form a network between each other that need not be constrained by the physical world. The responsibilities of a P2P overlay network includes providing peers the ability to join and leave the network and find other peers.

@@

The purpose of peer to peer networks (and more broadly, networks in general) is to share data. In the P2P networks that focus not just on sharing data, but also storing, locating, and possibly retrieving data (like ours), the data shares the same virtual address space as the peers. That is, structured overlay networks give the data themselves valid addresses. For example, in our system, the data being backed up is also given a 20 byte "address". This is done to determine which peer is responsible for a given piece of data. Specifically, a peer is responsible for all data in the address space that the peer is closest to. Aside: the size of the address space for which a peer is responsible decreases as more peers join the network, which allows these types of networks to easily scale in size.

To join a structured overlay network, a peer must know at least one other peer already in the network. This peer is often called the \textit{bootstrap peer}. The joining peer will contact the bootstrap peer to find its proper place in the topology. More details about finding the proper place in the topology are given below, in the explanation of how to find other peers.

@@

Leaving a structured overlay network can be as simple as not responding to any messages. Some specifications leave this functionality undefined, perhaps because peers may simply shutdown instead of gracefully quitting. Such overlay networks handle node departures as node failures. This is how our system behaves.

To find other peers, a peer in a structured overlay network will consult its routing table. This table is initially populated upon joining the network, and is kept up to date as peers join and leave. When a joining peer first contacts a bootstrap peer, it is unlikely that the bootstrap peer has the joining peer's direct neighbors in its routing table. This is because the routing table is designed to include many peers that are close, and only a few peers that are far. A joining peer, by asking the bootstrap node, will learn of the peers in the bootstrap peer's routing table that the joining node is close to. The joining peer will then contact these peers, ask for the same information, and learn of peers it is even closer to. This process repeats (usually in $O(\log n)$ time) until the joining node learns its direct neighbors \cite{p2pSurvey}.
% TODO explain O(logn)
% TODO can we relate this to our system somehow? If not, why are we talking about it? Maybe we should simplify... Though it would be unfortunate to not utilize this research into structured overlays we've done

@@

\subsection{Fairness}
Another aspect of P2P backup systems is fairness between peers. Peers need to be storing an amount of data proportional to the amount that they are backing up to other peers. This can be challenging, because untrustworthy peers can attempt to game the system to obtain free backup storage. For example, consider a simple system where each peer must offer store the same amount of data that they back up on other peers. A peer could claim that all of the storage space that it should be offering is being used by other peers, while in reality the peer has not taken in any peer's data. It is clear that a mechanism needs to be built into the system that ensures that peers store the data of each other. PeerStore handles this via a combined trading and challenge system. When a peer wants to create a backup of its data, it finds a peer to trade data with. In its request to backup data, the peer will also advertise that it is willing to store a certain amount of data. The peer that receives the request will then need to accept or reject it, and if the request is accepted the receiver will also need to request a certain amount of storage. If the original requestor accepts the offer, then the two become trading partners and will send each other data \cite{PeerStore}.

\begin{figure}
  \centering
  %\includegraphics[scale=0.75]{PeerStoreTrading}
  \caption{Trading protocol of PeerStore}
\end{figure}

While this ensures that peers will agree to take data in proportion to the amount that they want to backup, it still does not guarantee that peers hold onto that data. For example, a peer could form a trading agreement with another peer, exchange data, and promptly discard the received data. To combat this, peers can periodically challenge peers to prove that they are actually storing data. This is the approach that PeerStore takes. A challenge in PeerStore consists of a randomly-generated number followed by a specified range of bytes in the file being backed up. The node holding the peer's data must take the bytes in the specified range, concatenate the random number with that data, take the hash of it, and return the result to the peer. If the correct answer is returned, then the node has passed the challenge. If the node does not respond or gives the wrong answer, the challenge has been failed. By choosing a random subset of bytes and creating a random number to include in the hash, the node cannot calculate the hash ahead of time and only store that; the whole file must be stored in order to be able to successfully complete all possible challenges.

\subsection{Churn and Data Migration} \label{sec:churn}
With each storage node being regular users, it is often the case that nodes can be offline. \textit{Node churn} refers to fact that peers will continuously enter and leave the P2P network \cite{StorageSearchP2PNetworks}. The designer of a P2P backup system must take this into consideration if the system is to provide reliable access to backed up data. This has been accomplished via data migration \cite{pStore,PeerStore}, where the availability of files is periodically checked and replicas are made when the nodes being used for backup are not reliably available. Other data migration algorithms have been designed specifically to work in high churn networks \cite{StorageSearchP2PNetworks}. In this case, data replicas are periodically passed to different groups of nodes, which means that newer nodes have a chance to store data. This is beneficial because data gets evenly distributed throughout the overlay network, so any given node is not more important to the network than any others with regard to the amount of data it is storing.

The number of replicas to maintain will have an impact on the performance of the system. If the replica count is too low, the probability of recovery the entire file is lowered. If the replica count is too high, then more storage space and bandwidth will be wasted. If the system requires a peer to offer a proportional amount of storage space to the amount that it is using on other nodes, the peer will also have to offer a lot more space than what it is using locally for its own data.

\subsection{Cryptography} \label{sec:crypto}
% TODO: the figure for convergent encryption might be nice here
Since data is being stored on several untrusted peers, each user's data need to be encrypted before backup. A technique used in pStore and PeerStore is convergent encryption \cite{pStore, PeerStore}. To use this encryption mechanism, the hash of the data to encrypt is calculated, and then the hash is used as the symmetric key to encrypt the data. The main benefit of using convergent encryption is that the same data will be encrypted the same way each time, so identical blocks in the network can be shared between peers in their encrypted form.

\begin{figure}
  \centering
  %\includegraphics[scale=0.75]{ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore}
\end{figure}

While this is a convenient optimization, it allows for users who have access to the unencrypted version of a file to confirm whether or not that file is being stored, since no more information is needed to encrypt or decrypt a given file.
%\subsection{Case Study: pStore}
%\subsection{Case Study: PeerStore}
\section{Distributed Hash Tables} \label{sec:dht}

@@

\chapter{Implementation}
\section{Languages and Tools}
\section{System Design}
\subsection{Overview}
\subsection{}
\subsection{DHT}

@@

\bibitem{dropboxusers} https://www.dropbox.com/news/company-info
\bibitem{pStore} pStore
\bibitem{PeerStore} PeerStore
\bibitem{StorageSearchP2PNetworks} Storage and Search in P2P Networks
\bibitem{p2pSurvey} A survey on content-centric technologies for the current Internet: CDN and P2P solutions
\bibitem{bittorrentDHT} http://www.bittorrent.org/beps/bep\_0005.html
\bibitem{btsynctech} http://www.bittorrent.com/sync/technology
\bibitem{btsyncuserguide} http://btsync.s3-website-us-east-1.amazonaws.com/BitTorrentSyncUserGuide.pdf
\bibitem{btsyncapi} http://www.bittorrent.com/sync/developers/api
\bibitem{bittorrentProtocol} http://www.bittorrent.org/beps/bep\_0003.html

>>>>

%%\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{alltt}
\usepackage{color}
\usepackage{fancyvrb} 
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{url}
%\usepackage[autostyle]{csquotes} 
%\usepackage[toc,page]{appendix}

@@

\begin{document}
\begin{center}
\thispagestyle{empty}
% TODO install \textsc font
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\newpage
\end{center}

@@

In the client-server model, servers are expected to be always available and accessible. Peer to peer networks do not have this luxury. Instead, they experience the constant fluctuation of peers joining and leaving the network. This is called \textit{churn} \cite{StorageSearchP2PNetworks}. Churn in P2P networks complicates the communication between peers because the sought peer may be offline. This is especially a problem in P2P backup systems, for example, when trying to recover data from a peer who is not online. The strategies for dealing with churn, and many other P2P-specific complications, heavily depend on the type of overlay network a P2P application uses. We discuss churn in more detail in section \ref{sec:churn}.

@@

An overlay network is a network built on top of another network. That is, an overlay network consists of nodes with neighbors who are not physically connected, but instead use the underlying network to establish a virtual connection. They need not be constrained by the limitations of the physical world. Consider a trivial example where there are three computers connected to the Internet, but located in different parts of the world. These three computers could form an overlay network by agreeing to be virtual neighbors to each other. If their overlay network was used in a P2P application, then, when connecting to each other, they will use the underlying network, the Internet, to actually send messages. The P2P application does not need to know about the underlying network that is used to route messages, it only needs to know that there is an overlay network that defines the direct connections between peers.

Peer to peer applications use overlay networks as their foundation.\footnote{Overlay networks are used for many purposes, not just P2P applications. Note, however, in discussing overlay networks, we will refer to their participants as \textit{peers}.}  The P2P overlay network provides peers the ability to join and leave the network and find other peers with whom to communicate.

@@

The purpose of peer to peer networks (and more broadly, networks in general) is to share data. In the P2P networks that focus not just on sharing data, but also storing, locating, and possibly retrieving data (like ours), the data shares the same virtual address space as the peers. That is, these structured overlay networks give the data themselves valid addresses. For example, in our system, the data being backed up is also given a 20 byte "address". This is done to determine which peer is responsible for a given piece of data. Specifically, a peer is responsible for all data in the address space that the peer is closest to. It is for this reason that peer address assignment should be evenly distributed, to keep the distribution of work equal. Aside: the average size of the address space for which a peer is responsible decreases as more peers join the network, which allows these types of networks to easily scale in size.

To join a structured overlay network, a peer must know at least one other peer already in the network. This peer is often called the \textit{bootstrap peer}. The joining peer will contact the bootstrap peer to find its proper place in the topology. We address this in more detail in the explanation of how to find other peers.

@@

Leaving a structured overlay network can be as simple as not responding to any messages. Our system, like many others, leaves this functionality undefined. Node failures due to hard shutdowns will not follow the protocol to leave the network, so it is easiest to treat all node departures as node failures.

To find other peers, a peer in a structured overlay network will consult its routing table. This table is initially populated upon joining the network, and is kept up to date as peers join and leave. When a joining peer first contacts a bootstrap peer, it is unlikely that the bootstrap peer has the joining peer's direct neighbors in its routing table. This is because the routing table is designed to include many peers that are close, and only a few peers that are far\footnote{This is to provide an efficient, $O(\log n)$ time complexity. This means that given n nodes, it will only take $c \times \log n$ steps, which is good performance.} A joining peer, by asking the bootstrap node, will learn of the peers in the bootstrap peer's routing table that the joining node is close to. The joining peer will then contact these peers, ask for the same information, and learn of peers it is even closer to. This process repeats until the joining node learns its direct neighbors \cite{p2pSurvey}.

@@

\subsection{Fairness}
Another aspect of P2P backup systems is fairness between peers. Peers need to be storing an amount of data proportional to the amount that they are backing up to other peers. This can be challenging to enforce, because untrustworthy peers can attempt to game the system to obtain free backup storage. For example, consider a simple system where each peer must offer store the same amount of data that they back up on other peers. A peer could claim that all of the storage space that it should be offering is being used by other peers, while in reality the peer has not taken in any other peer's data. It is clear that a mechanism needs to be built into the system that ensures that peers store the data of each other.

PeerStore handles this via a combined trading and challenge system. When a peer wants to create a backup of its data, it finds a peer to trade data with. In its request to backup data, the peer will also advertise that it is willing to store a certain amount of data for its trading partner. The peer that receives the request will then need to accept or reject it, and if the request is accepted, the acceptor will also need to request a certain amount of storage in return. If the original requestor accepts this offer, then the two become trading partners and will send each other data. This process is illustrated in figure \ref{fig:peerStoreTrading} \cite{PeerStore}.

\begin{figure} \label{fig:peerStoreTrading}
  \centering
  \includegraphics[scale=0.75]{PeerStoreTrading}
  \caption{Trading protocol of PeerStore}
\end{figure}

While this ensures that peers will agree to take data in proportion to the amount that they want to backup, it still does not guarantee that peers hold onto that data. For example, a peer could form a trading agreement with another peer, exchange data, and promptly discard the received data. To combat this, peers can periodically challenge peers to prove that they are actually storing data. This is the approach that we, like PeerStore, propose. We provide peers the ability to ask each other to prove they are storing the data that they are supposed to be storing. A peer proves itself by replying with a value that could only be derived if they were, indeed, safekeeping the data. See \ref{sec:challenge} for more details about our proposed challenge mechanism.

\subsection{Churn and Data Migration} \label{sec:churn}
With each storage node being regular users, it is often the case that nodes can be offline. Recall that churn refers to fact that peers will continuously enter and leave the P2P network. The designer of a P2P backup system must take this into consideration if the system is to provide reliable access to backed up data. This has been accomplished via data migration \cite{pStore,PeerStore}, where the availability of files is periodically checked and replicas are made when the nodes being used for backup are not reliably available. Other data migration algorithms have been designed specifically to work in high churn networks \cite{StorageSearchP2PNetworks}. In this case, data replicas are periodically redistributed to different groups of nodes, which means that newer nodes have a chance to store data. This is beneficial because data stays evenly distributed throughout the overlay network, preventing any single node from being responsible for too much data, regardless of when they joined\footnote{It is important to note this strategy is designed to work in an unstructured overlay network, not a structured one. Structured networks need not redistribute like this because the evenly distributed address assignment provides the even data distribution.}

The number of replicas to maintain will have an impact on the performance of the system. If the replica count is too low, the probability of recovery the entire file is lowered. If the replica count is too high, then storage space and bandwidth will be needlessly used. Furthermore, if the system requires a peer to offer a proportional amount of storage space to the amount that it is using on other nodes, the peer will also have to offer a lot more space than what it is using locally for its own data. % TODO this last sentence might go better in the fairness section

\subsection{Cryptography} \label{sec:crypto}
Crytography is the study of techniques for secure communication in the presence of third parties \cite{cryptoDef}. It is heavily relied upon in the computer world to protect secrets, verify identities, and ensure data hasn't been tampered with. Our system uses cryptography to \textit{encrypt} users' data before replicating it to stranger's computers. Encryption is the process obfuscating text into a state such that it cannot be understood without reversing the process. There are several types of encryption proccesses, each with its respective \textit{decryption} process. Decryption refers to reversing the encryption so that encrypted message becomes readable again.

In \textit{symmetric key cryptography}, the encryption and decryption processes use identical \textit{keys}. The key is the secret parameter to the mathematical functions that prevents anyone else from decrypting the data. Anyone who has the key for a given message can decrypt it. Using a different key to encrypt the same data will result in a different output.

Encryption in P2P backup systems is necessary to prevent the peers who are storing backups of data from reading the data they are backing up. A system in which backups were unencrypted would be much less popular with users, because they would lose privacy by using it. The terminology associated with this is \textit{confidentiality}, meaning that data that is supposed to be confidential, remains confidential.

\textit{Hashing} is the process of applying a \textit{hash function} to some arbitrary-length message, resulting in a fixed sized \textit{hash}. A good hash function will uniformly distribute the hashes it produces. \textit{Cryptographic hash functions} are hash functions that are suitable for use in cryptography. They possess mathematical properties that allow for confidentiality to be provided, among other properties.

Convergent encryption is a technique used in pStore and PeerStore to ensure confidentiality \cite{pStore, PeerStore}. To use this encryption mechanism, the hash of the unencrypted data is calculated, and then the hash is used as the symmetric key to encrypt the data. The main benefit of using convergent encryption is that the same data will be encrypted the same way each time, so identical blocks in the network can be shared between peers in their encrypted form, without the need for a complicated key exchange protocol. Convergent encryption is illustrated in figure \ref{fig:convergent}.

\begin{figure} \label{fig:convergent}
  \centering
  \includegraphics[scale=0.75]{ConvergentEncryption}
  \caption{Convergent Encryption from PeerStore}
\end{figure}

While this is a convenient optimization, it has the following security weakness: any user who has access to the unencrypted version of a file can confirm whether or not that file is being stored in the system. This encroaches on users' privacy.

\section{Distributed Hash Tables} \label{sec:dht}

@@

\chapter{Implementation}
\section{Languages and Tools}
\section{System Design}
\subsection{Overview}
\subsection{}
\subsection{DHT}
\section{Ensuring Backup Safekeeping} \label{sec:challenge}
% This was copied from the explanation of challenge mechanisms in the background info. We will use it down here instead
%A challenge in PeerStore consists of a randomly-generated number followed by a specified range of bytes in the file being backed up. The node holding the peer's data must take the bytes in the specified range, concatenate the random number with that data, take the hash of it, and return the result to the peer. If the correct answer is returned, then the node has passed the challenge. If the node does not respond or gives the wrong answer, the challenge has been failed. By choosing a random subset of bytes and creating a random number to include in the hash, the node cannot calculate the hash ahead of time and only store that; the whole file must be stored in order to be able to successfully complete all possible challenges.

@@

\bibitem{dropboxusers} https://www.dropbox.com/news/company-info
\bibitem{pStore} pStore
\bibitem{PeerStore} PeerStore
\bibitem{StorageSearchP2PNetworks} Storage and Search in P2P Networks
\bibitem{p2pSurvey} A survey on content-centric technologies for the current Internet: CDN and P2P solutions
\bibitem{bittorrentDHT} http://www.bittorrent.org/beps/bep\_0005.html
\bibitem{btsynctech} http://www.bittorrent.com/sync/technology
\bibitem{btsyncuserguide} http://btsync.s3-website-us-east-1.amazonaws.com/BitTorrentSyncUserGuide.pdf
\bibitem{btsyncapi} http://www.bittorrent.com/sync/developers/api
\bibitem{bittorrentProtocol} http://www.bittorrent.org/beps/bep\_0003.html
\bibitem{cryptoDef} Rivest, Ronald L. (1990). "Cryptology". In J. Van Leeuwen. Handbook of Theoretical Computer Science 1. Elsevier.
